---
title: "Berkeley R User Group"
subtitle: "Dates with R"
author: "Dan Murphy"
date: "July 21, 2015"
output: 
    slidy_presentation
---

# Date vs Date-Time

> - CRAN's TimeSeries View: https://cran.r-project.org/web/views/TimeSeries.html

>> - lubridate: https://github.com/hadley/lubridate

>>> - "The methods we use with date-times must be robust to time zones, leap days, daylight savings times, and other time related quirks, and R lacks these capabilities in some situations."

> - R's Two Main Date Classes are **POSIX\*t** (Date-Time) and **Date** (Date-only)

>> - Two embodiments of POSIX*t

>>> - POSIXct

>>> - POSIXlt

>>> - Search ISO 8601:2004

<ul class="incremental">
<li>
```{r}
(ct <- as.POSIXct("2015-01-31 18:00"))
(lt <- as.POSIXlt("2015-01-31 18:00"))
(dt <- as.Date("2015-01-31 18:00"))
```
</li>

***

## What are these objects?

<ul class="incremental">
<li>
```{r}
class(ct)
class(lt)
class(dt)
```
</li>

## What's under the hood?

<li>
```{r}
mode(ct)
mode(lt)
mode(dt)
```
</li>
<li>
```{r}
as.numeric(ct)
unlist(lt)
as.numeric(dt)
```
</li>
</ul>

> * POSIXct = the number of seconds since 1/1/1970 in timezone "UTC" (aka, GMT)

> * Date = the number of days since 1/1/1970

<ul class="incremental">
<li>
```{r}
ct0 <- as.POSIXct("1970-01-01")
as.numeric(ct0)
dt0 <- as.Date("1970-01-01")
as.numeric(dt0)
```
</li>
</ul>

# Timezones are problematic

> * POSIX*t's have a timezone

<ul class="incremental">
<li>
```{r}
ct0
```
</li>
</ul>

> * Date's have __no__ timezone

<ul class="incremental">
<li>
```{r}
dt0
```
</li>
</ul>

> * Use 'tz' argument to specify a POSIX*t timezone

<ul class="incremental">
<li>
```{r}
ctUTC <- as.POSIXct("1970-01-01", tz = "UTC")
ctUTC
as.numeric(ctUTC)

# Extra credit: why did as.numeric(ct0) = 28800?
# Hint: 28800 = 8 * 60 * 60
```
</li>
</ul>

> * If 'tz' not specified, _local timezone_ is assumed

>> * ct and lt treat differently

<ul class="incremental">
<li>
```{r}
ct <- as.POSIXct("2015-07-21")
ct
attr(ct, "tzone")
lt <- as.POSIXlt("2015-07-21")
lt
lt$zone
```
</li>
</ul>

> * If timezone is important, be explicit

>> * Abbreviations __do not__ work

>>> * See https://en.wikipedia.org/wiki/List_of_time_zone_abbreviations

>> * See OlsonNames() for list of valid timezones on your OS

<ul class="incremental">
<li>
```{r}
Sys.timezone()
ct <- as.POSIXlt("1970-01-01", tz = "America/Los_Angeles")
ct
attr(ct, "tzone")
lt <- as.POSIXlt("1970-01-01", tz = "America/Los_Angeles")
lt
lt$zone
# Abbreviations don't work when specifying timezone
suppressWarnings(as.POSIXlt("1970-01-01", tz = "PST"))
# There were 11 warnings (use warnings() to see them)
```
</li>
</ul>

> * Recommendation: if timezone info not needed, use Date class for simplification

> * Many databases' date-time fields are based on the ISO standard (POSIXct),
so be careful if timezone is unspecified in the date field

# Date Arithmetic

> * A Year = 365 Days about 75% of the time

>> * 2016 is the next leap year

> * A Day always equals 60 x 60 x 24 = 86400 seconds in R,
even the days that have leap seconds
(e.g., June 30, 2015 at 23:59:60)

> * A Month is not 30, 31, 28, 29 days

> * What is one month from January 31, 2015?

<ul class="incremental">
<li>
```{r}
seq(from = as.Date("2015-01-31"), length = 2, by = "months")
```
</li>
</ul>

>> * Per R Help: Using "month" first advances the month without changing the day: 
if this results in an invalid day of the month,
it is counted forward into the next month

<ul class="incremental">
<li>
```{r}
seq(from=as.Date("2015-01-03"), length = 3, by = "months")
```
</li>

> * For accountants, end of business (EOB), end of month (EOM), 
end of year (EOY) are key moments in time

# mondate package
Keep track of dates in terms of months

* http://cran.r-project.org/package=mondate

* Damien Laker,
"Time Calculations for Annualizing Returns: the Need for Standardization,"
*The Journal of Performance Measurement*, 2008

# Examples

<ul class="incremental">
<li>
```{r}
x <- as.Date("2015-07-31")
x + 365
```
</li>
<li>
```{r}
suppressPackageStartupMessages(library(mondate))
y <- mondate("2015-07-31")
y + 12
```
</li>



# International vs. American Character Representation of Dates

## Month Day Year

# Sources of Information

- http://blog.revolutionanalytics.com/2009/06/converting-time-zones.html
- Bullet 3

## lubridate

http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/

https://github.com/hadley/lubridate/blob/master/NEWS

> * v1.3.0. treats math with month and year Periods more consistently. If adding 
or subtracting n months would result in a non-existent date, lubridate will 
return an NA instead of a day in the following month or year. For example, 
`ymd("2013-01-31") + months(1)` will return `NA` instead of `2013-03-04` 
as in v1.2.0. `ymd("2012-02-29") + years(1)` will also return an `NA`. This 
rule change helps ensure that date + timespan - timespan = date (or NA). If 
you'd prefer that such arithmetic just returns the last day of the resulting 
month, see `%m+%` and `%m-%`.
