---
title: "Berkeley R User Group"
subtitle: "Dates with R"
author: "Dan Murphy"
date: "July 21, 2015"
output: 
    slidy_presentation
---

# Date vs Date-Time

> - CRAN's TimeSeries View: https://cran.r-project.org/web/views/TimeSeries.html

>> - lubridate: https://github.com/hadley/lubridate

>>> - "The methods we use with date-times must be robust to time zones, leap days, daylight savings times, and other time related quirks, and R lacks these capabilities in some situations."

> - R's Two Main Date Classes are **POSIX\*t** (Date-Time) and **Date** (Date-only)

>> - Two embodiments of POSIX*t

>>> - POSIXct

>>> - POSIXlt

>>> - Search ISO 8601:2004

<ul class="incremental">
<li>
```{r}
(ct <- as.POSIXct("2015-01-31 18:00"))
(lt <- as.POSIXlt("2015-01-31 18:00"))
(dt <- as.Date("2015-01-31 18:00"))
```
</li>
</ul>
> * What's under the hood?

# What's under the hood?

<ul class="incremental">
<li>
```{r}
mode(ct)
mode(lt)
mode(dt)
```
</li>
<li>
```{r}
attributes(ct)
attributes(lt)
attributes(dt)
```
</li>
<li>
```{r}
as.numeric(ct)
unlist(lt)
as.numeric(dt)
```
</li>
</ul>

> * POSIXct = the number of seconds since 1970-01-01 00:00:00 UTC
(Universal Time Coordinated, aka GMT)

> * Date = the number of days since 1970-01-01

<ul class="incremental">
<li>
```{r}
ct0 <- as.POSIXct("1970-01-01")
as.numeric(ct0)
dt0 <- as.Date("1970-01-01")
as.numeric(dt0)
```
</li>

* 0, okay, but 28800? Hmmm
</ul>

# Timezones can be problematic

> * ?timezones

>> * "Time zones are a system-specific topic"

>> * "Unfortunately there are many system-specific errors in the implementations"

> * POSIX*t's have a timezone

<ul class="incremental">
<li>
```{r}
ct0
```
</li>
</ul>

> * Date's have __no__ timezone

<ul class="incremental">
<li>
```{r}
dt0
```
</li>
</ul>

> * Use 'tz' argument to specify a POSIX*t timezone

<ul class="incremental">
<li>
```{r}
ctUTC <- as.POSIXct("1970-01-01", tz = "UTC")
ctUTC
as.numeric(ctUTC)

# Extra credit: why did as.numeric(ct0) = 28800?
# Hint: 28800 = 8 * 60 * 60
```
</li>
</ul>

> * If 'tz' not specified, _local timezone_ is assumed

>> * ct and lt treat differently

<ul class="incremental">
<li>
```{r}
ct <- as.POSIXct("2015-07-21")
ct
attr(ct, "tzone")
lt <- as.POSIXlt("2015-07-21")
lt
lt$zone
```
</li>
</ul>

> * If timezone is important, be explicit with the 'tz' argument

>> * Abbreviations __do not__ work for 'tz'

>>> * See https://en.wikipedia.org/wiki/List_of_time_zone_abbreviations

>> * See OlsonNames() for list of valid 'tz' arguments

<ul class="incremental">
<li>
```{r}
Sys.timezone()
ct <- as.POSIXlt("1970-01-01", tz = "America/Los_Angeles")
ct
attr(ct, "tzone")
lt <- as.POSIXlt("1970-01-01", tz = "America/Los_Angeles")
lt
lt$zone
# Abbreviations don't work when specifying timezone
suppressWarnings(as.POSIXlt("1970-01-01", tz = "PST"))
# There were 11 warnings (use warnings() to see them)
#Warning messages:
#1: In strptime(xx, f <- "%Y-%m-%d %H:%M:%OS", tz = tz) : unknown timezone 'PST'
```
</li>
</ul>

> * Recommendation: if timezone not needed, use Date class for simplicity

> * Warning: Many databases' date-time fields are based on the ISO standard (POSIXct),
so __be careful if timezone is unspecified in the database__
because R will assume your local timezone

# Conversion from Date to DateTime

<ul class="incremental">
<li>
```{r}
dt0
as.POSIXct(dt0)
as.POSIXlt(dt0)
```
</li>
</ul>

> * __GRR!!!__

> * Recommendations when converting from Date to POSIX*t
-- e.g., before storing the numeric value back to a database

>> * Be agnostic as to timezone

>> * Therefore, assume timezone is UTC 

>> * When converting from Date to POSIX*t, __use lt__

>> * Can convert to ct later

<ul class="incremental">
<li>
```{r}
as.POSIXct(as.POSIXlt(dt0))
unclass(as.POSIXct(as.POSIXlt(dt0)))
```
</li>
</ul>

>> * Of course, the information on this page is moot 
if the database stores dates as character strings

# Date, DateTime Arithmetic

> * Adding or subtracting units to/from a DateTime object works best with integers

>> * number of seconds for POSIX*t

>> * number of days for Date

<ul class="incremental">
<li>
```{r}
ct0 + 3600
dt0 + 31
```
</li>
</ul>

> * 'difftime' object

>> * Represents the interval between two instances of time

>> * Can be created by subtracting two datetime objects

<ul class="incremental">
<li>
```{r}
as.POSIXct("1970-01-02") - ct0
as.Date("1970-01-02") - dt0
```
</li>
</ul>

>> * Or by creating a difftime object

<ul class="incremental">
<li>
```{r}
u <- as.difftime(1, units = "days")
u
ct0 + u
dt0 + u
```
</li>
</ul>

> * Allowable units are "auto", "secs", "mins", "hours",
"days", and "weeks"

>> * Missing "months" and "years"

> * "months" _is_ available with 'seq' for datetime objects 
(argument 'by')

<ul class="incremental">
<li>
```{r}
seq(from = dt0, length = 2, by = "months")
```
</li>
</ul>

> * What is one month from the last day of January?

<ul class="incremental">
<li>
```{r}
seq(from = as.Date("1970-01-31"), length = 2, by = "months")
seq(from = as.Date("1970-01-03"), length = 3, by = "months")
```
</li>
</ul>

>> * Per ?seq.POSIXt: 
Using "month" first advances the month without changing the day: 
if this results in an invalid day of the month,
it is counted forward into the next month

> * For accountants, end of business (EOB), end of month (EOM), 
end of year (EOY) are key moments in time

# mondate package

Keeps track of dates in terms of months

* http://cran.r-project.org/package=mondate

* Damien Laker,
"Time Calculations for Annualizing Returns: the Need for Standardization,"
*The Journal of Performance Measurement*, 2008

    * Proposed that dates be recorded as a percent of the month that has transpired
    
* Like POSIXct, the underlying representation of a mondate object 
is a "numeric"

    * reprresents the number of months that have transpired
since the end of the year 1999
    
# Examples

<ul class="incremental">
<li>
```{r}
x <- as.Date("2015-07-31")
x + 365
```
</li>
<li>
```{r}
library(mondate)
y <- mondate("2015-07-31")
y + 12
y + 1:12
```
</li>
</ul>

> * mondate can handle dates in US format

<ul class="incremental">
<li>
```{r}
y <- mondate("7/31/2015")
y + 1:12
as.Date(y + 1:12)
```
</li>
</ul>

> * Can be useful when dates stored in csv file are in US format
(e.g., saved from Excel)

# Thank you

Dan Murphy